Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-23 paper/main.tex
refs n cites

Both to divide genomes into training, validation and test sets, as well as to understand
performance, we needed a way to evaluate genome and annotation quality.
Metadata was generated for each genome, including: basic length and 
composition statistics \citep{gurevich2013quast}; k-mer content \citep{marccais2011fast};
conserved gene content of the genome, transcriptome, and proteome (\cite{simao2015busco};
file conversion with gffread from \cite{trapnell2012differential}; % BUSCO, files generated with gffread
and finally simple counts of genic features from the gff files. In particular, 
we considered the following features of the metadata when judging the quality of a genome and it's 
annotation: assembly contiguity (N50 / genome size); retention in the proteome 
of detected conserved orthogroups (BUSCOs) that were identified in the genome; and
the presence of UTRs and alternative splicing as estimated by the gff feature-counts.
Finally this metadata was combined with researcher knowledge of model species to guide genome
selection as necessary.

@@

We downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{} % supp phylogenies maybe?
within our generalizable range (i.e. excluding algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal
 data; the bottom row is concerned with the plants. Yellow dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all genomes that were not used during training of Helixer, which is 
equivalent to all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. See Section 4 for the 
Supplements shows the effects of overlapping for each individual species.}
\end{figure*}

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S1 and S3. We do not, for example, consistently predict very well on the 
algae as well as the non-avian reptiles. While one of the the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

AUGUSTUS outperforms us on some of the smallest genomes, but falls off much more 
drastically as the genomes get larger. The difference in prediction quality is 
especially strong for mammals, which tend to have big genomes with very long genes 
as well as the largest plants. Figures S2 and S4 display the comparison to AUGUSTUS 
by phylogenetic position.

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack a single precise biological site \citep{},
or noise in the reference, which we observed relative to the RNAseq data \ref{}. % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (b and c,hSlA2 supp). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (hSlA2, hSlA3, lSlA3). Notably, in one of the examples ()
the RNAseq shows evidence of alternative splicing; and in another () Helixer's
prediction falls between that of the reference and Augustus.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{sec:generalization}\end{supsec}
\begin{supsec}\label{sec:hyperparameters}\end{supsec}
\begin{supsec}\label{sec:longer}\end{supsec}
\begin{supsec}\label{sec:overlapping}\end{supsec}
\begin{supsec}\label{sec:training_data}\end{supsec}
\begin{supsec}\label{sec:dcnn}\end{supsec}
\begin{supsec}\label{sec:rnaseq_eval}\end{supsec}

>>>>

Both to divide genomes into training, validation and test sets, as well as to understand
performance, we needed a way to evaluate genome and annotation quality.
Metadata was generated for each genome, including: basic length and 
composition statistics \citep{gurevich2013quast}; k-mer content \citep{marccais2011fast};
conserved gene content of the genome, transcriptome, and proteome (\cite{simao2015busco};
file conversion with gffread from \cite{trapnell2012differential}; 
and finally simple counts of genic features from the gff files. In particular, 
we considered the following features of the metadata when judging the quality of a genome and it's 
annotation: assembly contiguity (N50 / genome size); retention in the proteome 
of detected conserved orthogroups (BUSCOs) that were identified in the genome; and
the presence of UTRs and alternative splicing as estimated by the gff feature-counts.
Finally this metadata was combined with researcher knowledge of model species to guide genome
selection as necessary.

@@

We downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{supfig:rna_select} % subgenic F1 sorted by Helixer new plots
within our generalizable range (i.e. excluding the outgroups algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal
 data; the bottom row is concerned with the plants. Orange dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all genomes that were not used during training of Helixer, which is 
equivalent to all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. 
See Section \ref{supsec:overlapping} of the 
Supplements for the effects of overlapping in individual species.}
\end{figure*}

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S\ref{supfig:phylogeny_f1_animals} and S\ref{supfig:phylogeny_f1_plants}. 
We do not, for example, consistently predict very well on the 
algae nor on the non-avian reptiles. While one of the the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

AUGUSTUS outperforms us on some of the smallest genomes, but falls off much more 
drastically as the genomes get larger. The difference in prediction quality is 
especially strong for mammals, which tend to have big genomes with very long genes 
as well as the largest plants. Figures S\ref{supfig:phylogeny_differences_animals} 
and S\ref{supfig:phylogeny_differences_plants} display the comparison to AUGUSTUS 
by phylogenetic position.

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer a), neither b), the reference, but not Helixer c),
and Helixer, but not the reference d).

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome},  % todo, citation for termination site as well
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c, S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (S\ref{supfig:cov_example_02}d))
the RNAseq shows evidence of alternative splicing; 
and in another (S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and Augustus.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_03}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{sec:generalization}\end{supsec}
\begin{supsec}\label{sec:hyperparameters}\end{supsec}
\begin{supsec}\label{sec:longer}\end{supsec}
\begin{supsec}\label{sec:overlapping}\end{supsec}
\begin{supsec}\label{sec:training_data}\end{supsec}
\begin{supsec}\label{sec:dcnn}\end{supsec}
\begin{supsec}\label{sec:rnaseq_eval}\end{supsec}
